Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-03-26 ident_manu.Rnw
calib results text done

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{cgem}[CGEM]{Coastal General Ecosystem Model}
\acrodef{do}[O$_2$]{dissolved oxygen}
\acrodef{dom}[DOM]{dissolved organic matter}
\acrodef{gom}[GOM]{Gulf of Mexico}
\acrodef{lcs}[LCS]{Louisiana continental shelf}
\acrodef{marb}[MARB]{Mississippi-Atchafalaya River Basin}
\acrodef{pom}[POM]{particulate organic matter}
\acrodef{zerod}[0-D]{zero-dimensional}

@@

% R dependencies
<<include = F, cache = F>>=
source('R/funcs.r')
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ggrepel)
library(english)
library(purrr)
library(grid)
library(lubridate)
# library(raster)
library(rgeos)
library(ggpolypath)
library(readxl)
library(ggmap)
library(ggsn)
@

@@

Selected parameter subsets from the seven heuristics were used to calibrate the \ac{zerod} model.  This analysis demonstrated that different parameter sets can produce model output with differing precision and calibration effort given 1) differences in variable sensitivity to parameters in each subset, 2) differences in the parameters selected from different categories (heuristic difference), and 3) differences in identifiability values.  Generally, this analysis was a proof of concept that the selection heuristics can lead to an `optimal' model by maximizing the balance between sensitivity and identifiability.

@@

Each parameter subset was calibrated to most closely match the model output to the observed time series.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This opimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within +/- 50\% of the default parameter values.  While we recognize that `actual' parameter values could possibly be outside of this range, constraining the values to those tested herein was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in precision as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the root-mean squared error between the observed and modelled data.

@@

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values.  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.
<<>>=
NULL
@

@@

library(tidyverse)
library(lubridate)
source("R/funcs.R")

load(file = 'data/tocal_all.RData')
load(file = 'data/cals.RData')

# starting rmse
strt <- cals[[1]]$mat[1,] %>% 
  .[length(.)]
# 72.13955

levs <- c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton', 'top', 'cattop')
labs <- c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton', 'Independent of category', 'Equally within category')
caldat <- lapply(cals, function(x) c(x$value, nrow(x$mat))) %>% 
  do.call('rbind', .) %>% 
  data.frame %>% 
  rownames_to_column %>% 
  rename(
    heurist = rowname, 
    rmse = X1, 
    its = X2
  ) %>% 
  left_join(colerr, by = 'heurist') %>% 
  select(heurist, coll, nprm, avee, its, rmse) %>% 
  mutate(
    heurist = factor(heurist, levels = levs, labels = labs),
    coll = ifelse(is.na(coll), 1, coll),
    coll = sapply(coll, scinot),
    nprm = sapply(nprm, scinot),
    its = sapply(its, scinot),
    rmse = sapply(rmse, scinot), 
    avee = sapply(avee, scinot)
    ) %>% 
  arrange(heurist)

# pairs(caldat[, -1])
totab <- caldat
cap_val <- 'Results of model calibration to observed \\ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\\gamma$) and parameters of each subset are from \\cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.'
colnms <- c('$\\gamma$', '$n$ parameters', 'mean L1', 'Iterations', '$RMSE$')

latex( 
  totab[, -1],
  file = '',
  rowlabel = 'Selection heuristic',
  caption = cap_val,
  caption.loc = 'top',
  size = 'scriptsize',
  label = 'tab:calib',
  rowname = totab$heurist,
  rgroup = c('1', '2', '3'),
  n.rgroup = c(5, 1, 1),
  colheads = colnms
  )

>>>>

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{cgem}[CGEM]{Coastal General Ecosystem Model}
\acrodef{do}[O$_2$]{dissolved oxygen}
\acrodef{dom}[DOM]{dissolved organic matter}
\acrodef{gom}[GOM]{Gulf of Mexico}
\acrodef{lcs}[LCS]{Louisiana continental shelf}
\acrodef{marb}[MARB]{Mississippi-Atchafalaya River Basin}
\acrodef{pom}[POM]{particulate organic matter}
\acrodef{rmse}[\textit{RMSE}]{root mean squared error}
\acrodef{zerod}[0-D]{zero-dimensional}

@@

% R dependencies
<<include = F, cache = F>>=
source('R/funcs.r')
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ggrepel)
library(english)
library(purrr)
library(grid)
library(lubridate)
# library(raster)
library(rgeos)
library(ggpolypath)
library(readxl)
library(ggmap)
library(ggsn)
library(tibble)
@

@@

Selected parameter subsets from the seven heuristics were used to calibrate the \ac{zerod} model.  This analysis demonstrated that different parameter sets can produce model output with differing precision and calibration effort given
\begin{inparaenum}[1\upshape)]
\item differences in variable sensitivity to parameters in each subset, 
\item differences in the parameters selected from different categories (heuristic difference), and 
\item differences in identifiability values.
\end{inparaenum}
Generally, this analysis was a proof of concept that the selection heuristics can lead to an `optimal' model by maximizing the balance between sensitivity and identifiability.

@@

Each parameter subset was calibrated to most closely match the model output to the observed time series.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This opimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within +/- 50\% of the default parameter values.  While we recognize that `actual' parameter values could possibly be outside of this range, constraining the values to those tested herein was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in precision as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the \ac{rmse} between the observed and modelled data.

@@

<<>>=
load(file = 'data/tocal_all.RData')
load(file = 'data/cals.RData')

# starting rmse
# 72.13955

levs <- c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton', 'top', 'cattop')
labs <- c('Optics', 'Organic Matter', 'Phytoplankton', 'Temperature', 'Zooplankton', 'Independent of category', 'Equally within category')
caldat <- lapply(cals, function(x) c(x$value, nrow(x$mat))) %>% 
  do.call('rbind', .) %>% 
  data.frame %>% 
  rownames_to_column %>% 
  rename(
    heurist = rowname, 
    rmse = X1, 
    its = X2
  ) %>% 
  left_join(colerr, by = 'heurist') %>% 
  select(heurist, coll, nprm, avee, its, rmse) %>% 
  mutate(
    heurist = factor(heurist, levels = levs, labels = labs)
  )

# pairs(caldat[, -1])
temp <- filter(caldat, heurist %in% 'Temperature')
optics <- filter(caldat, heurist %in% 'Optics')
top <- filter(caldat, heurist %in% 'Independent of category')
cattop <- filter(caldat, heurist %in% 'Equally within category')
@

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values (\cref{tab:calib}).  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.  As expected, the number of required iterations to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only \Sexpr{toeng(temp$its)[[1]]} iterations were required to optimize the single parameter selected for the temperature category, whereas \Sexpr{top$its} iterations were required for the \Sexpr{toeng(top$nprm)[[1]]} parameters selected independent of category.  A slight positive assocation between collinearity and number of iterations was also observed, such that computational effort  to identify optimal values for a parameter set inreased with lower identifiability (higher $\gamma$).  Of the seven paramter subsets, selection independent of category and equally within each category produced the lowest \ac{rmse} values (\Sexpr{scinot(top$rmse)}, \Sexpr{scinot(cattop$rmse)}, respectively), whereas the optimal values for paramters in the Optics category had the lowest reduction in \ac{rmse} (\Sexpr{scinot(optics$rmse)}).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 (\Sexpr{scinot(optics$avee)}), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 (\Sexpr{scinot(top$avee)}, \cref{tab:calib}).

@@

strt <- cals[[1]]$mat[1,] %>% 
  .[length(.)] %>% 
  scinot

totab <- caldat %>% 
  mutate(
    coll = ifelse(is.na(coll), 1, coll),
    coll = sapply(coll, scinot),
    nprm = sapply(nprm, scinot),
    its = sapply(its, scinot),
    rmse = sapply(rmse, scinot), 
    avee = sapply(avee, scinot)
    ) %>% 
  arrange(heurist)

cap_val <- 'Results of model calibration to observed \\ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\\gamma$) and parameters of each subset are from \\cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.'
foot.val <- '\\footnotesize *Starting \\ac{rmse} $= 72.14$'
colnms <- c('$\\gamma$', '$n$ parameters', 'mean L1', 'Iterations', '$RMSE$*')

latex( 
  totab[, -1],
  file = '',
  rowlabel = 'Selection heuristic',
  caption = cap_val,
  caption.loc = 'top',
  size = 'scriptsize',
  label = 'tab:calib',
  rowname = totab$heurist,
  rgroup = c('1', '2', '3'),
  insert.bottom = foot.val, 
  n.rgroup = c(5, 1, 1),
  colheads = colnms
  )

### 2017-03-26 ident_manu.tex
calib results text done

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Sun Mar 26 15:18:56 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/bf9e2162ec0d9ba7edf4ddc5b7ba5c1019142186}{bf9e2162ec0d9ba7edf4ddc5b7ba5c1019142186}}

@@

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values.  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.
NULL

@@

% calibration table, do
%latex.default(totab[, -1], file = "", rowlabel = "Selection heuristic",     caption = cap_val, caption.loc = "top", size = "scriptsize",     label = "tab:calib", rowname = totab$heurist, rgroup = c("1",         "2", "3"), n.rgroup = c(5, 1, 1), colheads = colnms)%
\begin{table}[!tbp]
{\scriptsize
\caption{Results of model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\gamma$) and parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.\label{tab:calib}} 
\begin{center}
\begin{tabular}{llllll}
\hline\hline
\multicolumn{1}{l}{Selection heuristic}&\multicolumn{1}{c}{$\gamma$}&\multicolumn{1}{c}{$n$ parameters}&\multicolumn{1}{c}{mean L1}&\multicolumn{1}{c}{Iterations}&\multicolumn{1}{c}{$RMSE$}\tabularnewline
\hline
{\bfseries 1}&&&&&\tabularnewline
~~Optics&$3.87$&$3$&$2.81\times 10^{-4}$&$14$&$72.14$\tabularnewline
~~Organic Matter&$8.61$&$5$&$3.18\times 10^{-3}$&$154$&$68.7$\tabularnewline
~~Phytoplankton&$3.97$&$7$&$0.02$&$270$&$68.72$\tabularnewline
~~Temperature&$1$&$1$&$0.02$&$6$&$71.76$\tabularnewline
~~Zooplankton&$5.2$&$4$&$0.03$&$99$&$71.22$\tabularnewline
\hline
{\bfseries 2}&&&&&\tabularnewline
~~Independent of category&$6.81$&$8$&$0.03$&$850$&$67.05$\tabularnewline
\hline
{\bfseries 3}&&&&&\tabularnewline
~~Equally within category&$5.22$&$7$&$0.03$&$690$&$67.67$\tabularnewline
\hline
\end{tabular}\end{center}}
\end{table}

>>>>

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Sun Mar 26 15:32:37 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/b41a6df79d21b187457db88f058313b9b430416a}{b41a6df79d21b187457db88f058313b9b430416a}}

@@

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values (\cref{tab:calib}).  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.  As expected, the number of required iterations to minimize \ac{rmse} of observed \ac{do} and model output varied in proportion to the number of parameters in a subset.  For example, only six iterations were required to optimize the single parameter selected for the temperature category, whereas 850 iterations were required for the eight parameters selected independent of category.  A slight positive assocation between collinearity and number of iterations was also observed, such that computational effort  to identify optimal values for a parameter set inreased with lower identifiability (higher $\gamma$).  Of the seven paramter subsets, selection independent of category and equally within each category produced the lowest \ac{rmse} values ($67.05$, $67.67$, respectively), whereas the optimal values for paramters in the Optics category had the lowest reduction in \ac{rmse} ($72.14$).  Lower \ac{rmse} was associated with higher average sensitivity of a parameter subset.  Parameters selected only in the optics category had the highest \ac{rmse} and lowest average L1 ($2.81\times 10^{-4}$), whereas parameters selected independent of category had the lowest \ac{rmse} and highest average L1 ($0.03$, \cref{tab:calib}).

@@

% calibration table, do
%latex.default(totab[, -1], file = "", rowlabel = "Selection heuristic",     caption = cap_val, caption.loc = "top", size = "scriptsize",     label = "tab:calib", rowname = totab$heurist, rgroup = c("1",         "2", "3"), insert.bottom = foot.val, n.rgroup = c(5,         1, 1), colheads = colnms)%
\begin{table}[!tbp]
{\scriptsize
\caption{Results of model calibration to observed \ac{do} for each of seven parameter subsets selected by sensitivity within each parameter category (first), independent of category (second), and equally within each category (third). The collinearity values ($\gamma$) and parameters of each subset are from \cref{tab:heurist1,tab:heurist2,tab:heurist3}.  Mean L1 is the average of sensitivity values (\cref{l1}) for the subset.  $RMSE$ shows the final error comparing observed and model output after the specified number of iterations for optimization.\label{tab:calib}} 
\begin{center}
\begin{tabular}{llllll}
\hline\hline
\multicolumn{1}{l}{Selection heuristic}&\multicolumn{1}{c}{$\gamma$}&\multicolumn{1}{c}{$n$ parameters}&\multicolumn{1}{c}{mean L1}&\multicolumn{1}{c}{Iterations}&\multicolumn{1}{c}{$RMSE$*}\tabularnewline
\hline
{\bfseries 1}&&&&&\tabularnewline
~~Optics&$3.87$&$3$&$2.81\times 10^{-4}$&$14$&$72.14$\tabularnewline
~~Organic Matter&$8.61$&$5$&$3.18\times 10^{-3}$&$154$&$68.7$\tabularnewline
~~Phytoplankton&$3.97$&$7$&$0.02$&$270$&$68.72$\tabularnewline
~~Temperature&$1$&$1$&$0.02$&$6$&$71.76$\tabularnewline
~~Zooplankton&$5.2$&$4$&$0.03$&$99$&$71.22$\tabularnewline
\hline
{\bfseries 2}&&&&&\tabularnewline
~~Independent of category&$6.81$&$8$&$0.03$&$850$&$67.05$\tabularnewline
\hline
{\bfseries 3}&&&&&\tabularnewline
~~Equally within category&$5.22$&$7$&$0.03$&$690$&$67.67$\tabularnewline
\hline
\end{tabular}\end{center}}
\footnotesize *Starting \ac{rmse} $= 72.14$\end{table}
